RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-03-12 inst/doc/pubs/manuscript.Rmd
writing writing

@@

### Writing NeXML phylogenies
### Reading NeXML metadata

One of the most powerful features of the NeXML format is the ability add arbitrary metadata information using the RDFa annotation standard [cite W3C]().

Using the nexml and nexmlTree class for metadata.

@@

`RNeXML` also helps users add richer metadata when producing NeXML files. This occurs on several levels:

@@

@@

The `RNeXML` package performs two levels of validation on NeXML files
written from or read into R: validating the XML, and validating against
the current NeXML schema.

Several mechanisms are in place to evaluate the quality of the `RNeXML`
package functions themselves.  In addition to the examples from the
documentation and the automated package checks provided for the R system,
the `RNeXML` package includes a complete test suite using the testthat
package [Wickham 2011].

@@

[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/
[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html
[Chamberlain 2013]: http://dx.doi.org/10.12688/f1000research.2-191.v2
[@ouch]: http://cran.r-project.org/web/packages/ouch/
[@phylobase]: http://cran.r-project.org/web/packages/phylobase

>>>>

At the heart of the NeXML standard is the ability to encode arbitrary metadata in precisely specified, machine-readable formats that maintain backwards compabitibilty with any existing NeXML parser. NeXML uses the "Resource Description Framework in Annotations", or RDFa [W3C 2013], as described in [Vos et al 2011].

```{r } 
birds <- nexml_read("test.xml")
get_taxa(birds)
get_metadata(birds) 
```

@@

@@

Researchers cannot take advantage of NeXML metadata manipulation if there is no metadata there.  The burden of adding good metadata remains a significant barrier to increased sharing and reuse of scientific research [Tenopir et al. 2011].  `RNeXML` seeks to reduce the burden of adding high quality, machine-readable metadata to NeXML files. This occurs on several levels:

@@

- Extending the NeXML standard through metadata annotation.

No data standard can anticipate all the uses and extensions researchers may one day need. Researchers have repeatedly adapted the existing NEXUS file format to accomidate encoding new kinds of information, but in the process break compatibility between different flavors of NEXUS files, with few reliable methods to distinguish between these flavors.  NeXML provides a solution to this problem through the use of metadata annotation.

Here we illustrate this process using the example of stochastic character mapping [Huelsenbeck et al 2003]. A stochastic character map is simply an annotation of the branches on a phylogeny, assigning each section of each branch to a particular "state" (typically of a morphological characteristic).

[Bollback 2006] provides a widely used stand-alone software implementation of this method in the software `simmap`, which modified the standard Newick tree format to express this additional information. This can break compatibility with other software[^1], and creates a format that cannot be interpreted without additional information describing this convention.  By contrast, the NeXML extension is not only backwards compatible but contains a precise and machine-readable description of what it is encoding.

[^1]: By using the commenting mechanism of the Newick format, it is possible that other software that doesn't also use the comment mechanism for some other such purpose would be able to successfully parse the tree.  However there is no way to guarentee that this is the case or for the data format to describe its use.

[Revell 2011] describes the `phytools` package for R, which includes utilities for reading, manipulating, and writing `simmap` files.

NeXML annotation of an edge that switches from state `s2` to state `s1`:

```xml
<edge about="#e34" id="e34" source="n16" target="n35" length="0.205223">
        <meta xsi:type="nex:LiteralMeta" id="m262" property="simmap:reconstructions" datatype="xsd:string">
          <meta xsi:type="nex:LiteralMeta" id="m260" property="simmap:reconstruction" datatype="xsd:string">
            <meta xsi:type="nex:LiteralMeta" id="m261" property="simmap:char" datatype="xsd:string" content="cr1"/>
            <meta xsi:type="nex:LiteralMeta" id="m252" property="simmap:stateChange" datatype="xsd:string">
              <meta xsi:type="nex:LiteralMeta" id="m253" property="simmap:order" datatype="xsd:decimal" content="1"/>
              <meta xsi:type="nex:LiteralMeta" id="m254" property="simmap:length" datatype="xsd:decimal" content="0.202947550453246"/>
              <meta xsi:type="nex:LiteralMeta" id="m255" property="simmap:state" datatype="xsd:string" content="s2"/>
            </meta>
            <meta xsi:type="nex:LiteralMeta" id="m256" property="simmap:stateChange" datatype="xsd:string">
              <meta xsi:type="nex:LiteralMeta" id="m257" property="simmap:order" datatype="xsd:decimal" content="2"/>
              <meta xsi:type="nex:LiteralMeta" id="m258" property="simmap:length" datatype="xsd:decimal" content="0.00227544954675424"/>
              <meta xsi:type="nex:LiteralMeta" id="m259" property="simmap:state" datatype="xsd:string" content="s1"/>
            </meta>
          </meta>
        </meta>
      </edge>
```

[Revell 2011]: http://doi.org/10.1111/j.2041-210X.2011.00169.x
[Huelsenbeck et al 2003]: http://doi.org/10.1080/10635150390192780 "Stochastic Mapping of Morphological Characters"
[Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."

@@

The `RNeXML` package performs severak levels of validation on NeXML files
written from or read into R: validating the XML, and validating against
the current NeXML schema, and performing additional checks through the 
NeXML online validation tool for elements that cannot be enforced by
schema validation alone.  Validation serves to improve consistency in
data encoding which facilitates the use of NeXML across a wide variety
of platforms and software.  A lack of a consistent standard and validation
mechanism has continually plagued the current NEXUS standard and made 
interoperability difficult [Vos et al. 2011].

Several mechanisms are in place to evaluate the quality of the `RNeXML`
package functions themselves.  In addition to the examples from the
documentation and the automated package checks provided for the R system,
the `RNeXML` package includes a complete test suite using the `testthat`
package [Wickham 2011]. Package development on Github includes continuous 
integration through the `travis` platform, in which the full test suite
is run every time the code is changed.  This not only gives users confidence
that the development copy can be successfully installed, but should assist
other developers seeking to contribute to the package to test that their proposed 
changes have not negatively impacted any of the existing code base.

@@

[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/
[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html
[Chamberlain 2013]: http://dx.doi.org/10.12688/f1000research.2-191.v2
[@ouch]: http://cran.r-project.org/web/packages/ouch/
[@phylobase]: http://cran.r-project.org/web/packages/phylobase
[W3C 2013]: http://www.w3.org/TR/xhtml-rdfa-primer/ "RDFa 1.1 Primer - Second Edition: Rich Structured Data Markup for Web Documents"
[Tenopir et al. 2011]: http://doi.org/10.1371/journal.pone.0021101 "Data Sharing by Scientists: Practices and Perceptions"
